\documentclass[12pt]{article}
%% arXiv paper template by Flip Tanedo
%% last updated: Dec 2016


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%  THE USUAL PACKAGES  %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{nopageno}
\usepackage{enumerate}
\usepackage{parskip}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%  UNUSUAL PACKAGES        %%%%
%%%  Uncomment as necessary. %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% MATH AND PHYSICS SYMBOLS
%% ------------------------
%\usepackage{slashed}       % \slashed{k}
%\usepackage{mathrsfs}      % Weinberg-esque letters
%\usepackage{youngtab}	    % Young Tableaux
%\usepackage{pifont}        % check marks
\usepackage{bbm}           % \mathbbm{1} incomp. w/ XeLaTeX 
%\usepackage[normalem]{ulem} % for \sout


%% CONTENT FORMAT AND DESIGN (below for general formatting)
%% --------------------------------------------------------
\usepackage{lipsum}        % block of text (formatting test)
%\usepackage{color}         % \color{...}, colored text
%\usepackage{framed}        % boxed remarks
%\usepackage{subcaption}    % subfigures; subfig depreciated
%\usepackage{paralist}      % compactitem
%\usepackage{appendix}      % subappendices
%\usepackage{cite}          % group cites (conflict: collref)
%\usepackage{tocloft}       % Table of Contents	

%% TABLES IN LaTeX
%% ---------------
%\usepackage{booktabs}      % professional tables
%\usepackage{nicefrac}      % fractions in tables,
%\usepackage{multirow}      % multirow elements in a table
%\usepackage{arydshln} 	    % dashed lines in arrays

%% Other Packages and Notes
%% ------------------------
%\usepackage[font=small]{caption} % caption font is small



\renewcommand{\thesection}{}
\renewcommand{\thesubsection}{\arabic{subsection}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%  PAGE FORMATTING and (RE)NEW COMMANDS  %%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage[margin=2cm]{geometry}   % reasonable margins

\graphicspath{{figures/}}	        % set directory for figures

% for capitalized things
\newcommand{\acro}[1]{\textsc{\MakeLowercase{#1}}}    

\numberwithin{equation}{subsection}    % set equation numbering
\renewcommand{\tilde}{\widetilde}   % tilde over characters
\renewcommand{\vec}[1]{\mathbf{#1}} % vectors are boldface

\newcommand{\dbar}{d\mkern-6mu\mathchar'26}    % for d/2pi
\newcommand{\ket}[1]{\left|#1\right\rangle}    % <#1|
\newcommand{\bra}[1]{\left\langle#1\right|}    % |#1>
\newcommand{\Xmark}{\text{\sffamily X}}        % cross out


\let\olditemize\itemize
\renewcommand{\itemize}{
  \olditemize
  \setlength{\itemsep}{1pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}


% Commands for temporary comments
\newcommand{\comment}[2]{\textcolor{red}{[\textbf{#1} #2]}}
\newcommand{\flip}[1]{{\color{red} [\textbf{Flip}: {#1}]}}
\newcommand{\email}[1]{\texttt{\href{mailto:#1}{#1}}}

\newenvironment{institutions}[1][2em]{\begin{list}{}{\setlength\leftmargin{#1}\setlength\rightmargin{#1}}\item[]}{\end{list}}


\usepackage{fancyhdr}		% to put preprint number



%%%%%%%%%%%%%%%%%%%
%%%  HYPERREF  %%%%
%%%%%%%%%%%%%%%%%%%

%% This package has to be at the end; can lead to conflicts
\usepackage{microtype}
\usepackage[
	colorlinks=true,
	citecolor=black,
	linkcolor=black,
	urlcolor=green!50!black,
	hypertexnames=false]{hyperref}



%%%%%%%%%%%%%%%%%%%%%
%%%  TITLE DATA  %%%%
%%%%%%%%%%%%%%%%%%%%%

\begin{document}


\begin{center}

    {\Large \textsc{Homework 5b:} 
    \textbf{Extra Dimensions} (Optional)}
    
\end{center}

\vskip .4cm

\noindent
\begin{tabular*}{\textwidth}{rlcrll}
	\textsc{Course:}& Physics 231, \emph{Methods of Theoretical Physics} (2018)
	&
%	\hspace{1.2cm}
	&
	\\
	\textsc{Instructor:}& Professor Flip Tanedo (\email{flip.tanedo@ucr.edu})
	&
	%\hfill
	&
	& 
	\\
	\textsc{Due by:}& \emph{Friday}, Dec 7
	&
	%\hfill
	&
	%	
\end{tabular*}


This homework is encouraged, but optional.


\subsection{The Wave Equation in (3+1) Dimensions}
\label{problem:wave:eq:3d}

We solve for the Green's function of wave equation for electromagnetism: 
\begin{align}
	\left[\frac{1}{c^2}\frac{\partial^2}{\partial t^2} - \frac{\partial^2}{\partial \vec{r}^2}\right] 
	\varphi(\vec{r},t) &= \rho(\vec{r},t)
	&
	\left[\frac{1}{c^2}\frac{\partial^2}{\partial t^2} - \frac{\partial^2}{\partial \vec{r}^2}\right]
	\vec A(\vec{r},t) &= \vec j(\vec{r},t) \ .
	\label{eq:EM:wave}
\end{align}
We make the speed of light, $c$, explicit and have chosen \textbf{Lorentz gauge} and are working in vacuum, $\varepsilon_0 = \mu_0 = 1$. 
%
You should recognize all the physical quantities here. 
% 
Use spacetime coordinates such that $x = (ct,\vec r)$. Define the 4-vector potential $A_\mu(x) = \left(\varphi(x),\vec A(x)\right)$, where $x$ is a point in Minkowski space. Let $G(x,x')$ be the Green's function for each component of $A_\mu$. 
%
This means that the potential is given by
\begin{align}
	A_\mu(x) = \int d^4x \, G(x,x') \, j_\mu(x') \ .
\end{align}
Make sure you're comfortable with the indices\footnote{Be able to answer the following questions for yourself: Why doesn't $G(x,x')$ have indices? Why shouldn't it be a 4-component object or a matrix?}. 

In comparison to what we did in lecture, we'll keep factors of $c$ explicit and will write out the space and time components separately rather than writing $\partial^2 = \partial_t^2 -\partial_x^2 -\partial_y^2 -\partial_z^2$.




\subsubsection{Fourier Transform}

The Green's function equation for each component is
\begin{align}
	\left[\frac{1}{c^2}\frac{\partial^2}{\partial t^2} - \frac{\partial^2}{\partial \vec{r}^2}\right] G(x,x') = \delta^{(4)}(x-x') \equiv \delta(x-x')\delta(y-y')\delta(z-z')\delta(t-t') \ .
	\label{eq:4D:define:G}
\end{align}
We write the Fourier transform in Minkowski space \emph{covariantly}:
\begin{align}
	\tilde G(k,x') &= \int d^4x \, e^{ik\cdot x} G(x,x')
	&
	G(x,x') &= \int \dbar^4k \, e^{-ik\cdot x} \tilde G(k,x') \ ,
\end{align}
where we use the notation $\dbar = d/2\pi$. Further, $k = (E/c,\vec k)$ and we recall that $k\cdot x \equiv k_\mu x^\mu = Et - k_x x - k_y y - k_z z$. 
%\textcolor{green!50!black}{[\textsc{Update 11/27}: you may find it useful to think of this as $k\cdot x \equiv (E/c)(ct) - \vec k\cdot \vec x$.} 
Why is $k_0$ identified $E$? If you're not sure, wait until later and see what $E$ is fixed to be. 
%\textcolor{green!50!black}{[\textsc{Update 11/27}: note that $\dbar^4k = (dE/c)d^3\vec{k} (2\pi)^{-4}$]}

Solve for $\tilde G(k,x')$. It should look very similar, perhaps with factors of $c$ as required by dimensional analysis.  Write $G(x,x')$ as a 4D integral over $\tilde G(k,x')$.

\textsc{Hint:} It may be useful to recall that $\delta^{(4)}(x)$ has the following Fourier transform: 
\begin{align}
	\delta^{(4)}(x-x') &= \int \dbar^4 k e^{-ik\cdot (x-x')} \ .
\end{align}
%\textcolor{green!50!black}{[\textsc{Update 11/27}: Note that $\delta^{(4)}(x) = \delta(ct)\delta^{3}(\vec{x})$. Similarly, $\delta^{(4)}(k) = \delta(E/c)\delta^{3}(\vec{k})$. This is consistent with the definition of the volume element and the dot product. }

\textsc{Answer:} You will find
\begin{align}
	G(x,x') = \int \dbar^4k \, \frac{c^2}{c^2 \vec k^2 - E^2} e^{-ik\cdot (x-x')} \ .
	\label{eq:G:ft:1}
\end{align} 
Observe that $G(x,x') = G(x-x')$, as we expect from spacetime translation invariance.



\subsubsection{Angular integrals in hyper-cylindrical coordinates}

For convenience, write $y \equiv x-x' = (cu,\mathbf{s})$. This means that $u$ is a shifted time coordinate and $\vec s$ is a shifted spatial 3-vector. Note that $d^4x = d^4 y$. In order to integrate (\ref{eq:G:ft:1}), use 4D `hyper-cylindrical coordinates' over $k$ where time/energy is treated linearly and the space/momentum directions are treated in 3D spherical coordinates, $|\vec k|$, $\cos \theta$, and $\varphi$. Recall that $\theta$ is the azimuthal angle with respect to the $k_z$-axis. The volume element in these coordinates is
\begin{align}
	d^4 k = dE \, d^3\vec k = |\vec k|^2 \,dE \,  d|\vec k| \, d\cos \theta \, d\varphi \ .
\end{align}
%\textcolor{green!50!black}{[\textsc{Update 11/27}: Corrected $dE/c$ factor in above equation.]}

Since we integrate over all values of $k^\mu$, we are free to align our axes however we want. A particularly convenient choice is to align $k_z$ to be aligned with $\vec s$. In this case, $\vec k \cdot \vec s = ks \cos \theta$, where it should be understood that $k$ and $s$ are magnitudes of the spatial $3$-vectors\footnote{That is, from now on I write $k = |\vec k|$. There should not be any ambiguity with the 4-vector, $k^2 = E^2/c^2 - \vec k^2$, which no longer shows up in our expressions.}.
%
Then we have
\begin{align}
	-ik\cdot y = i k s \cos \theta  - i E u \ .
\end{align}
Perform the angular integrals in (\ref{eq:G:ft:1}).  Show that this gives
\begin{align}
	G(y) = \frac{1}{4\pi^3 s} \int_0^\infty k\,\sin ks 
	\left(
	 \int_{-\infty}^\infty \frac{c^2}{c^2k^2 - E^2} e^{-iE u} dE
	\right) \, dk \ .
	\label{eq:G:cauchy:P:problem}
\end{align}
%\textcolor{green!50!black}{[\textsc{Update 11/27}: Corrected $dE/c$ factor in above equation.]}





\subsubsection{Practical Pole Pushing for Poor Physics People}

The $dE$ integral in parenthesis in (\ref{eq:G:cauchy:P:problem}) looks very familiar. We evaluate it using a contour integral. Identify the location of the two poles in the complex $E$ plane.
%
You're already an expert on this integral.
%
You have a choice of contours: $\mathcal C_+$ in the upper half plane and $\bar{\mathcal C}_-$ in the lower half plane. These are parameterized by:
\begin{align}
	\mathcal C_+ &: \quad z = Re^{i\theta} \qquad  0 < \theta < \pi \\
	-{\mathcal C}_- &: \quad z = Re^{i\theta} \qquad \pi < \theta < 2 \pi \ ,
\end{align}
where in each case we take $R\to \infty$.
Comment on which contour one should choose depending on the sign of $u\equiv t-t'$.

Now we have to pick a prescription for how to navigate the poles. There are two physically-motivated choices: one can push the poles into the upper half plane and into the area enclosed by $\mathcal C_+$, or one can push the poles into the lower half plane and into the area enclosed by $-{\mathcal C}_-$. Which one corresponds to the causal case? Perform the $dE$ integral for this case.


\subsubsection{The Green's Function}

You now have $G(y)$ as a single integral over $dk$. Go ahead and perform this integral. 

\textsc{Hints}: Use, once again, the fact that $\int e^{ikx} dk = 2\pi \delta(x)$. The most straightforward calculation involves writing the solution to the $dE$ integral as a sum of exponentials rather than a sine. 

\textsc{Intermediate step}: It is useful to show that
\begin{align}
	\int_{-\infty}^\infty \frac{-e^{-i(E/c)(cu)}}{[(E/c)- k - i\varepsilon][(E/c)+k - i\varepsilon]}  dE = 
	\frac{-i\pi c}{k} \left(e^{icku}-e^{-icku}\right) \ .
\end{align}
It may be useful to define a variable $\mathcal E = E/c$.]
%
Another useful tip is to use
\begin{align}
	\int_0^\infty \left(e^{ikX} + e^{-ikX}\right) dk = 
	\int_{-\infty}^\infty e^{ikX} dk=
	2\pi \delta(X)
	 \ .
\end{align}


\textsc{Intermediate Step}: Writing $s = |\vec s|$, you should find:
\begin{align}
	G^{(r)}(y) &=
	\frac{c}{4\pi r}
	\left\{
	\begin{array}{ll}
%		\frac{c}{4\pi r} 
%		\left[
		\delta(s-cu) - \delta(s+cu)
%		\right]
		&
		\quad\text{if $u=t-t' \geq 0$}
		\\
		0 & \quad\text{if $u=t-t' <0$} \ .
	\end{array}
	\right.
	\label{eq:Gr:Gs}
\end{align}
Then observe that the radial coordinate $s\geq 0$ so that one of the $\delta$ functions in each expression will always be zero. 

\textsc{Answer}:
Taking $x' = 0$ so that $u=t$ and $s = r$, the final expression is:
\begin{align}
	G^{(r)}(x) = \frac{c}{4\pi r} \delta(r-ct) \Theta(t) \ .
	\label{eq:3d:G}
\end{align}
The physical interpretation is that the Green's function solves the electromagnetic wave equation for a `blip' impulse at $t=0$ and $\vec{r} = 0$. 




 





\subsection{The Wave Equation in Flatland}

There's a fantastic novella written in 1884 called \emph{Flatland: A Romance of Many Dimensions}\footnote{\url{https://en.wikipedia.org/wiki/Flatland}; see also \url{http://www.geom.uiuc.edu/~banchoff/Flatland/}}. It is a story about a 2+1 dimensional universe that doubled as a social commentary. In this problem, we explore electromagnetic waves in such a universe. 

\textsc{Note}: We walked through most of this problem in Lecture 20, feel free to follow along with the lecture notes.

%\subsection{Green's Function: set up}

We follow the same procedure as the previous problem.
\begin{align}
	\left[\frac{1}{c^2}\frac{\partial^2}{\partial t^2} - \frac{\partial^2}{\partial \vec{r}^2}\right] 
	\varphi(\vec{r},t) &= \rho(\vec{r},t)
	&
	\left[\frac{1}{c^2}\frac{\partial^2}{\partial t^2} - \frac{\partial^2}{\partial \vec{r}^2}\right]
	\vec A(\vec{r},t) &= \vec j(\vec{r},t) \ .
	\label{eq:EM:wave}
\end{align}
The Green's function for each component is:
\begin{align}
	\left[\frac{1}{c^2}\frac{\partial^2}{\partial t^2} - \frac{\partial^2}{\partial \vec{r}^2}\right] G(x,x') = \delta^{(3)}(x-x') \equiv \delta(x-x')\delta(y-y')\delta(t-t') \ .
	\label{eq:HO:in:3d}
\end{align}

\subsubsection{Fourier Transform}

Solve for the Green's function using a Fourier transform,
\begin{align}
	\tilde G(k,x') &= \int d^3x \, e^{ik\cdot x} G(x,x')
	&
	G(x,x') &= \int \dbar^3k \, e^{-ik\cdot x} \tilde G(k,x') \ ,
\end{align}
where continue to use the notation $\dbar = d/2\pi$. Further, $k = (E/c,\vec k)$ and we recall that $k\cdot x \equiv k_\mu x^\mu = Et - k_x x - k_y y$. Also note the $\delta$ function in 3-space: $\delta^{(3)}(x-x') = \int \dbar^3 k e^{-ik\cdot (x-x')}$ \ .

\textsc{Answer:} It should be no surprise to you at all that you end up with the following:
\begin{align}
	G(x,x') = \int \dbar^3k \, \frac{c^2}{c^2 \vec k^2 - E^2} e^{-ik\cdot (x-x')} \ .
	\label{eq:G:ft}
\end{align} 


\subsubsection{Something familiar}

For convenience, write $y \equiv x-x' = (cu,\mathbf{s})$. Check that it is still true that
\begin{align}
	-ik\cdot y = i k s \cos \theta  - i E u \ .
\end{align}
Unlike Problem~\ref{problem:wave:eq:3d}, however, our integration measure is
\begin{align}
	d^3 k = dE \, d^2\vec k = |\vec k| \,dE \,  d|\vec k| \, d \theta  \ .
\end{align}
\textsc{Warning:} At this stage, you have plenty of options for which integral to do first. Mathematically, it doesn't matter which order you do them---but some paths have more tedious integrals than others. I'm going to give hints for doing the integrals in the following order: $dE$, $d|\vec k|$, $d\theta$. 


Go ahead and perform the $dE$ integral for the case of the retarded Green's function. \textsc{Hint}: hey, haven't you done this \emph{exact} integral before? 

\textsc{Answer}:
\begin{align}
\int dE \frac{e^{-iEt}}{c|\vec k|^2 -E^2}
&=
\frac{-i\pi}{c|\vec k|} 
\left(
e^{ic|\vec k| t}
-
e^{-ic|\vec k| t}
\right) \ .
\label{eq:dE}
\end{align}

\textsc{Comment}: At this point, you might want to make a mental note of where $i\varepsilon$'s would be if we were keeping track of them.

\subsubsection{The $d|\vec k|$ integral}

Now perform the $d\theta$ integral. I suggest doing this the following way: write the result of (\ref{eq:dE}) into a sine function. Then use the relation
\begin{align}
	\int_0^\infty e^{iAk}\sin(ckt)\, dk = \frac{ct}{-A^2 + c^2t^2} \ .
\end{align}
\textsc{Remark:} This is a slightly mathematically dubious step which is okay if we judiciously kept track of the $i\varepsilon$. Specifically, this only holds when $t > r|\cos\theta|$. We'll be sloppy about this for now, we'll see the actual constraints come out more cleanly below.

\textsc{Answer}: You will find:
\begin{align}
	G(y) = \frac{1}{(2\pi)^2} \int d\theta \frac{ct}{c^2t^2 - r^2\cos^2\theta} \ .
\end{align}

\subsubsection{The angular integral}

A useful trick here is to use
\begin{align}
	\int_0^{2\pi} \frac{d\theta}{1-B\cos^2\theta} = 2\pi \sqrt{\frac{1}{1-B}} \ .
\end{align}
This only holds when $B<1$. 

\textsc{Answer}: You will find
\begin{align}
	G(y) = \frac{1}{2\pi} \frac{c}{\sqrt{c^2 t^2 - s^2}} \ ,
	\label{eq:2d:G:hardway}
\end{align}
with the caveat that $c^2t^2 > s^2$. Convince yourself that this makes sense from the point of view of causality.


\subsubsection{Dimensional Reduction}

% Now do it the Appel way.

The result (\ref{eq:2d:G:hardway}) solves the problem. 
%
There's actually a much nicer way to get the lower-dimensional Green's function from a higher-dimensional Green's function. 

Prove that 
\begin{align}
	G(x,y,t) = \int_{-\infty}^\infty G_{(3)}(x,y,z,t) \, dz \ ,
	\label{eq:G:from:G3}
\end{align}
where $G_{(3)}$ is the (3+1)-dimensional Green's function, (\ref{eq:3d:G}). In other words: one can derive the (2+1)-dimensional Green's function by integrating over the third dimension.

We prove this result without using the explicit form of $G_{(3)}(x,y,z,t)$, just the defining relation (\ref{eq:4D:define:G}). Write the right-hand side of (\ref{eq:G:from:G3}) as a Fourier transform, then do the $dz$ integral:
\begin{align}
	\int_{-\infty}^\infty dz G_{(3)}(x,y,z,t)dz
	&=
	\int \dbar^4k \,
	\left[dz\; e^{ikz} \tilde G_{(3)}(E,k_x, k_y, k_z)\right]
	 e^{-iEt} e^{i(k_xx + k_y y)} \ .
\end{align}
The $dz$ integral is just an inverse Fourier transform that gives $2\pi\delta(k_z)$. Perform the $k_z$ integral to write the right-hand side of (\ref{eq:G:from:G3}) in the form of a (2+1)-dimensional Fourier transform: $\int \dbar^3 k \, \tilde{\mathfrak{g}}e^{-iEt}e^{i(k_xx+k_yy)}$. 

Writing $\tilde{\mathfrak{g}}$ as the Fourier transform of the right-hand side of (\ref{eq:G:from:G3}), show that (\ref{eq:4D:define:G}) implies
\begin{align}
	\left(E^2 - k_x^2 - k_y^2\right) \tilde{\mathfrak{g}} = 1\ ,
\end{align}
thus proving that $\tilde{\mathfrak{g}}$ is the Fourier transform of the (2+1)-dimensional Green's function.

\subsubsection{Dimensional Reduction: explicit result}

For the retarded Green's function in (3+1)-dimensions, we found
\begin{align}
	G_{(3)}(\vec r,t) &= \frac{c}{4\pi r} \delta(r-ct)
	&
	\text{for } t&>0 \ ,
\end{align}
and zero for $t<0$. Integrate this expression over $dz$ to obtain the (2+1)-dimensional Green's function, $G(x,y,t)$,
\begin{align}
	G(x,y,t) &= \frac{c}{4\pi} \int_{-\infty}^\infty 
	\frac{1}{\sqrt{x^2+y^2 + z^2}} \delta\left(\sqrt{x^2+y^2 + z^2} - ct\right) \, dz \ .
\end{align}

\textsc{Hint:} you have one integral to do, but fortunately you have a $\delta$-function to do it. The only problem is that the $\delta$ function is of the form $\delta\left(f(z)\right)$. One way to solve this is to do a change of variables to $w = \sqrt{x^2+y^2+z^2}$ so that the $\delta$-function is of the form $\delta(w-ct)$. Doing this more generally gives a convenient rule,
\begin{align}
	\delta\left(f(z)\right) &= \sum_{z_i} \frac{1}{|f'(z_i)|} \delta(z-z_i)
	&
	z_i \text{ such that } f(z_i) = 0 \ .
\end{align}

\textsc{Answer:} The retarded Green's function is:
\begin{align}
	G(x,y,t)  &= 
	\left\{
	\begin{array}{ll}
		\frac{c}{2\pi} \frac{1}{\sqrt{c^2 t^2 - x^2 - y^2}}
		& \text{if } c^2t^2 > x^2 + y^2\\
		0 & \text{otherwise}
	\end{array}
	\right. \ .
\end{align}


\subsubsection{Discussion: The Odd Electrodynamics of Flatland}

Sketch $G(r,t)$ as a function of $r = \sqrt{x^2 + y^2}$ for two values of $t$. Compare this to the corresponding plot for $G_{(3)}(r,t)$. 

Think about how strange this lower-dimensional universe is! In our universe, if you used a flash camera in the dark, the each part of the room would light up for a brief instant corresponding to when the photons from the flash reaches each part of the room. In this lower-dimensional universe, $G(r,t)$ is telling us that the entire room remains illuminated for an \emph{infinite} amount of time with a rapidly decreasing (but never strictly zero) intensity.











\appendix
\section*{\Large Extra Credit}


These problems are not graded and are for your edification. You are strongly encouraged to explore and discuss these topics, especially if they are in a field of interest to you.

\subsection{Invariance of the Green's Function}

\subsubsection{Invariance of that funny $\delta$-function}

In Problem~\ref{problem:wave:eq:3d}, you might be concerned about is the covariance of the $\delta$-function with respect to Lorentz transformations. The Green's function $G(x)$ is supposed to be invariant under Lorentz transformations. However, the argument of $\delta(ct - r)$ does not appear to Lorentz invariant. As a reminder, under a Lorentz transformation $x^\mu \to \Lambda^\mu_{\phantom\mu\nu}x^\nu$ (this is matrix multiplication). However, the scalar quantity $x^2 = x_\mu x^\mu$ remains unchanged (invariant).

Show that the Green's function is indeed invariant.


\textsc{Hint:} Recall that the following rule for $\delta$-functions:
\begin{align}
	\delta\left(g(x)\right) &= \sum_{x_i} \frac{\delta(x-x_i)}{|g'(x_i)|}
	&
	g(x_i) = 0 \ .
\end{align}
Writing $x = (ct,\vec r)$, show that
\begin{align}
	\delta(x^2) = \frac{1}{2r} \left[
	\delta(r-ct) - \delta(r+ct)
	\right] \ .
\end{align}
Compare this to (\ref{eq:Gr:Gs}). 

\textsc{Answer}: You will find that 
\begin{align}
	G^{(r)}(x) = \frac{c}{2\pi} \delta(x^2) \Theta(t)
\end{align}
where $\Theta(t)$ is the Heaviside step function. Argue that even though $t=x^0$ is itself not invariant, $\Theta(t)$ is invariant under any \emph{physical} Lorentz transformation. 


\subsubsection{Discussion}
% Appel ex. 8.9
It turns out that when calculating matrix elements (amplitudes) for quantum mechanical scattering a relativistic particle of mass $m$ with four-momentum $p = (E,\vec p)$, you often end up with expressions in momentum space where an integrand $f(p)$ needs to be integrated over phase space in a way where the on-shell conditions are fixed:
\begin{align}
	p^2 &= E^2 - \vec p^2 = m^2 
	&
	E &>0
	\ .
\end{align}
In other words: you have an integral over the four directions in momentum space, but `physicality' imposes a constraint. You should be dreaming of Lagrange multipliers\footnote{We won't use Lagrange multipliers, but a generalization of this method where the $\delta$-function is promoted to a Lagrange multiplier shows up in gauge theory and is called the Fadeev-Popov procedure.}. 

What one ends up doing is writing down integrals of the form
\begin{align}
	\int d^4p \, \delta(p^2 - m^2) \Theta(E) f(p) \ .
\end{align}
This says: \emph{integrate over all of the phase space for $p^\mu$, but restrict it to positive energies and to 4-momenta that satisfy the $p^2=m^2$.}

Writing $E(\vec p) = \sqrt{\vec p^2 +m^2}$, show that
\begin{align}
	\int d^4p \, \delta(p^2 - m^2) \Theta(E) f(p) \ .
	& = 
	\int \frac{d^3\vec p}{2E(\vec p)} f(E(\vec p), \vec p) \ .
\end{align}
This funny-looking differential element $d^3\vec p/2E(\vec p)$ doesn't look Lorentz invariant, but we have derived that it is. Note that in the function, $f$, we fix $E$ to be the required value by `on-shell-ness.'




\subsection{Dimensional Regularization}

% e.g. Wikipedia, Cahill
% contour integral derivation
% do volumes of spheres in N dimensions
% Cahill p. 188

The Euler $\Gamma$ function is a meromorphic generalization of the factorial,
\begin{align}
	\Gamma(z) = \int_0^\infty x^{z-1}e^{-x} dx \ .
	\label{eq:Gamma}
\end{align}
For positive integers, $\Gamma(n) = (n-1)!$.
A fact that shows up occasionally in physics that area of the unit sphere in $d$ dimensions is
\begin{align}
	S_d = \frac{2\pi^{d/2}}{\Gamma(d/2)} \ .
	\label{eq:Sd}
\end{align}
There's always ambiguity (to me at least) when people say things like ``$d$-sphere'' ... is this a sphere embedded in $d$ dimensions whose area has units $L^{d-1}$, or is this the sphere embedded in $(d+1)$ dimension whose area has units $L^d$? What is the correct interpretation of (\ref{eq:Sd})? What is the area of the $d$-sphere of radius $r$? What is the [hyper]-volume enclosed by this sphere? (This is discussed in Cahill's textbook, around exercise 4.12.)

% Cahill ex 4.12


In physics, one often has to deal with quantifying and understanding apparent divergences in our calculations\footnote{These non-analytic results are the sign of our theory trying to tell us something.}. A systematic treatment of how to do this is usually buried deep in a course of quantum field theory, where there is a danger of confusing the physics from the mathematical techniques. In this problem we explore one such technique called \textbf{dimensional regularization} in a very simple electrostatic system. 

Consider the case of an infinite line  with constant charge density $\lambda$. (This should remind you of the dimensional reduction in Problem 3!) Pick coordinates such that the charged line is along the $y$-axis, and the observer is on the $x$-axis\footnote{By rotational symmetry $x$ may as well be the radial cylindrical coordinate, $\rho$.}. Observe that the potential (an integral over the Green's function!) at $x$ is
\begin{align}
	V(x) = \frac{\lambda}{4\pi} \int_{-\infty}^\infty \frac{dy}{\sqrt{x^2+y^2}} \ .
	\label{eq:Vx:1d}
\end{align}
Note that this integral is divergent\footnote{The expression is also scale invariant: $V(x) = V(kx)$. Think about what this means physically: an infinite line charge looks the same when you're close to it versus when you are far from it. This makes sense: you have no other length scale to compare to---thus by dimensional analysis, you have no basis to even use the words `close' or `far.'}. What does this mean and how do we make sense of quantities like $\vec E = -\nabla V$ in this case? As physicists, we deal with infinite quantities by \textbf{regulating} them: we \emph{quantify} the infinity by some parameter which allows us to compute finite physical quantities like $\vec E$ and check that those physical quantities are independent of the regulator.

Regulate the $dy$ integral (\ref{eq:Vx:1d}) by converting the one-dimensional integral into a ($1-2\epsilon$)-dimensional integral, where $\epsilon$ is a small number. Don't stop to ask what this means\footnote{\url{https://www.youtube.com/watch?v=ItV8utelYlc}. You can (and should) think about what this `slightly less than one dimension' means after you see the end result.}. This means we write:
\begin{align}
	V(x) = \frac{\lambda}{4\pi} \int_0^\infty dy\,d\Omega_{(1-2\epsilon)} \frac{y^{-2\epsilon}}{\mu^{-2\epsilon}}\frac{1}{\sqrt{x^2+y^2}} \ ,
\end{align}
where $d\Omega_{(1-2\epsilon)}$ is the differential area of the unit sphere in $(1-2\epsilon)$ dimensions, the factor of $y^{-2\epsilon}$ is inserted the give the right measure for a $(1-2\epsilon)$-dimensional volume, and some \emph{arbitrary} dimensionful parameter $\mu$ has been introduced to make units work out\footnote{I am using units such that $[V] = [\lambda]$. You can put in $\epsilon_0$ as appropriate, if you really wish.}. 
\begin{enumerate}[(a)]
\item Show that this integral evaluates to
\begin{align}
	V(x) = \frac{\lambda}{4\pi} \frac{\mu^{2\epsilon}}{x^{2\epsilon}} \frac{\Gamma(\epsilon)}{\pi^\epsilon} \ .
\end{align}
\item Confirm that as $\epsilon\to 0$, $V(x)$ becomes ill defined. \textsc{Hint}: Recall from $\Gamma(z+1) = z \Gamma(z)$ that $\Gamma(z)$ has a simple pole at the origin.
\item Calculate the electric field in the $x$ direction; $E_x = -\partial_xV(x)$ and show that
\begin{align}
	E_x = \frac{\lambda}{4\pi} \left[ \frac{2\epsilon \mu^{2\epsilon} \Gamma(\epsilon)}{\pi^\epsilon x^{1+2\epsilon}} \right] \ .
\end{align}
\item Take the $\epsilon \to 0$ limit and observe that $E_x = \lambda/2\pi x$, which is independent of the arbitrary scale $\mu$ and behaves the way that one expects. 
\end{enumerate}

\textsc{Hint:} Use the relation
\begin{align}
	\int_0^\infty \frac{y^{n-1}dy}{\sqrt{x^2+y^2}} = \frac{1}{2}\frac{\Gamma(\frac n2)\Gamma(\frac{1-n}2)}{\sqrt{\pi}}x^{n-1} \ .
\end{align}
This problem is worked out in \url{http://arxiv.org/abs/0812.3578}; you are strongly encouraged to read and think carefully about the entire article.







\end{document}